---
title: "Data Visualisation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    fig.width=6,
    fig.height=4
)
```

## First Steps

```{r mpg_1}
ggplot(data = mpg) + 
    geom_point(mapping = aes(x = displ, y = hwy))
```

### Exercises

1. Run `ggplot(data = mpg)`. What do you see?

```{r 2_1}
ggplot(data=mpg)
```
2. How many rows are in mpg? How many columns?

```{r 2_2}
dim(mpg)
```
3. What does the `drv` variable describe? Read the help for `?mpg` to find out.

the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd

4. Make a scatterplot of hwy vs cyl.

```{r 2_4}
mpg %>% 
    ggplot(aes(x = cyl, y = hwy)) +
    geom_point()
```

5. What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?

```{r 2_5}
mpg %>% 
    ggplot(aes(x = class, y = drv)) +
    geom_point()
```


Intersection of two factors. Contains no quantitive data.


## Aesthetic Mappings

1. What’s gone wrong with this code? Why are the points not blue?

```{r 3_1_1}
ggplot(data = mpg) + 
    geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

```{r 3_1_2}
ggplot(data = mpg) + 
    geom_point(mapping = aes(x = displ, y = hwy), colour = 'blue')
```
2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

```{r 3_2_1}
mpg %>% 
    map_lgl(is.factor)
```


```{r 3_2_2}
mpg
```
3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r 3_3}

```


4. What happens if you map the same variable to multiple aesthetics?

```{r 3_4}
mpg %>% 
    ggplot(aes(x = cyl, y = hwy, size = hwy, color = hwy)) +
    geom_point()
```

5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use `?geom_point`)

```{r 3_5}
# stroke modifies the width of the border
mpg %>% 
    ggplot(aes(x = cyl, y = hwy)) +
    geom_point(shape = 21, color = 'black', fill = 'white', stroke = 1)
```

6. What happens if you map an aesthetic to something other than a variable name, like `aes(colour = displ < 5)`? Note, you’ll also need to specify x and y.

```{r 3_6}
mpg %>% 
    ggplot(aes(x = cyl, y = hwy, colour = displ < 5)) +
    geom_point()
```

## Facets

Wrap

```{r facet_wrap, fig.width=8, fig.height=5}
ggplot(data = mpg) + 
    geom_point(mapping = aes(x = displ, y = hwy)) + 
    facet_wrap(~ class, nrow = 2)
```

Grid

```{r facet_grid, fig.width=8, fig.height=6}
ggplot(data = mpg) + 
    geom_point(mapping = aes(x = displ, y = hwy)) + 
    # rows ~ columns
    facet_grid(drv ~ cyl)
```

### Exercises

1. What happens if you facet on a continuous variable?

```{r 4_1, fig.width=8, fig.height=8}
mpg %>% 
    ggplot(aes(x = displ, y = hwy)) +
    geom_point() +
    facet_wrap(~cty)
```

2. What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?

```{r 4_2_1}
ggplot(data = mpg) + 
    geom_point(mapping = aes(x = drv, y = cyl))
```
The empty cells correspond the combinations of drv and cyl that have no records in common.

3. What plots does the following code make? What does `.` do?

```{r 4_3, fig.width=7, fig.height=5}
# rows
ggplot(data = mpg) + 
    geom_point(mapping = aes(x = displ, y = hwy)) +
    facet_grid(drv ~ .)

# columns
ggplot(data = mpg) + 
    geom_point(mapping = aes(x = displ, y = hwy)) +
    facet_grid(. ~ cyl)
```

4. Take the first faceted plot in this section:

```{r 4_4, fig.width=7, fig.height=5}
ggplot(data = mpg) + 
    geom_point(mapping = aes(x = displ, y = hwy)) +
    facet_wrap(~ class, nrow = 2)
```

What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?

5. Read `?facet_wrap`. What does `nrow` do? What does `ncol` do? What other options control the layout of the individual panels? Why doesn’t `facet_grid()` have nrow and ncol arguments?

6. When using `facet_grid()` you should usually put the variable with more unique levels in the columns. Why?

## Geometric Objects

### Exercises

1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

2. Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

```{r 5_1}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
    geom_point() + 
    geom_smooth(se = FALSE)
```


