---
title: "Relational Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    cache = FALSE
)

library(tidyverse)
library(nycflights13)
library(Lahman)
```

## nycflights13

- airlines lets you look up the full carrier name from its abbreviated code
- airports gives information about each airport, identified by the faa airport code
- planes gives information about each plane, identified by its tailnum
- weather gives the weather at each NYC airport for each hour

1. Imagine you wanted to draw (approximately) the route each plane flies from its origin to its destination. What variables would you need? What tables would you need to combine?


```{r ex_2_1_1}
flights %>% 
    # join origin to airports
    inner_join(airports, by=c('origin'='faa')) %>% 
    select(origin, name, lat, lon, dest) %>% 
    # join dest to airports
    inner_join(airports, by=c('dest'='faa'), suffix=c('_origin', '_dest')) %>% 
    select(matches('origin|dest'))
```



2. I forgot to draw the relationship between weather and airports. What is the relationship and how should it appear in the diagram?

```{r ex_2_1_2}
# airports with weather
airports %>% 
    semi_join(weather, by=c('faa'='origin'))
```

## Keys

1. Add a surrogate key to flights

```{r ex_3_1_1}
flights %>% 
    mutate(index=row_number()) %>% 
    select(index, everything())
```

2. Identify the keys in the following datasets

    1. `Lahman::Batting`
    1. `ggplot2::diamonds`


Use count aggregation:

    
```{r 3_1_2_1}

Batting %>% 
    count(playerID, yearID, stint, teamID) %>% 
    filter(n > 1)
```

```{r}
diamonds %>% 
    count(carat, cut, color, clarity, depth, table, price, x, y, z) %>% 
    filter(n > 1)
```


## Mutating Joins

1. Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:

```{r airports_quickmap}

airports %>%
    filter(lon > -140) %>% 
    # remove airports not in flights
    semi_join(flights, c("faa" = "dest")) %>%
    ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

```{r 4_1}
flights %>% 
    group_by(dest) %>% 
    summarise(avg_dep_time = mean(dep_time, na.rm = TRUE)) %>% 
    inner_join(airports, c('dest' = 'faa')) %>% 
    filter(lon > -140, !is.na(avg_dep_time)) %>% 
    select(lon, lat, avg_dep_time) %>% 
    ggplot(aes(x = lon, y = lat, size = avg_dep_time, color = avg_dep_time)) +
    borders('state') +
    guides(size = guide_none()) +
    labs(color='Average\ndeparture time') +
    geom_point() +
    coord_quickmap()
```

2. Add the location of the origin and destination (i.e. the lat and lon) to flights.

```{r 4_2}
flights %>% 
  left_join(airports, by = c('dest' = 'faa')) %>% 
  left_join(airports, by = c('origin' = 'faa'), suffix = c('.dest', '.origin')) %>% 
  select(dest, origin, contains('lat'), contains('lon')) 

```



3. Is there a relationship between the age of a plane and its delays?

Information about delays is in flights and the age is in planes

```{r 4_3}

flights %>% 
    # make tailnum the primary key
    group_by(tailnum) %>% 
    # get info about the average delays
    summarise(
        arrivals = mean(arr_delay, na.rm=T),
        departures = mean(dep_delay, na.rm=T)
    ) %>% 
    # long form
    pivot_longer(
        c(arrivals, departures),
        names_to='mode',
        values_to='delay',
        names_transform = list(mode=as.factor)
    ) %>% 
    # get info about the age
    left_join(planes, by='tailnum') %>% 
    select(tailnum, mode, delay, year) %>% 
    ggplot(aes(x = year, y = delay)) +
    geom_point(alpha=0.4) +
    geom_smooth(se = FALSE) +
    facet_wrap(~mode) +
    labs(x = 'Year of manufacture')
```



4. What weather conditions make it more likely to see a delay?


Join on year, month, day, hour, origin. Note that weather columns are observations, not variables and we want to map depature delay to weather conditions

```{r 4_4_1}
flights_weather <- flights %>% 
    left_join(weather, by = c('year', 'month', 'day', 'hour', 'origin')) %>% 
    # create a factor column where the levels are the conditions
    pivot_longer(temp:visib, names_to = 'condition', values_to = 'value') %>% 
    # consider only flights with a delay
    filter(!is.na(dep_delay)) %>% 
    # select the conditions and delay
    select(condition, value, dep_delay)
```


```{r 4_4_2, fig.width=12, fig.height=16}
flights_weather %>%  
    ggplot(aes(x = value, y = dep_delay)) +
    geom_hex() +
    labs(y = NULL) +
    facet_wrap(~condition, ncol = 3, scale = 'free_x')
```


5. What happened on June 13 2013? Display the spatial pattern of delays and then use Google to cross-reference with the weather.

```{r 4_5_1}
flights_june_6_2013_delays <- flights %>% 
    filter(year == 2013, month == 6, day == 13) %>% 
    group_by(dest) %>% 
    summarise(
        avg_arr_delay = mean(arr_delay, na.rm = TRUE),
        avg_dep_delay = mean(dep_delay, na.rm = TRUE)
    )
```



```{r 4_5_2}
flights_june_6_2013_delays %>% 
    left_join(airports, by = c('dest' = 'faa')) %>% 
    filter(lon > -140) %>% 
    ggplot(aes(x = lon, y = lat, size = avg_arr_delay, color = avg_arr_delay)) +
    borders('state') +
    geom_point(alpha = .8) +
    guides(size = guide_none()) +
    scale_color_continuous(low = 'yellow', high = 'red') +
    coord_quickmap()
```


## Filtering Joins
