---
title: "Strings"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE
)
library(tidyverse)
```

## String Basics

### Combining Strings

```{r combining_strings_1}
str_c("x", "y", sep = ", ")
str_c(
    c("a", "x"),
    c("b", "y"),
    sep = ", "
)
```

NAs are contagious

```{r combining_strings_2}
x <- c("abc", NA)
str_c("|-", x, "-|")
str_replace_na(x)

str_c("|-", str_replace_na(x), "-|")
```

```{r combining_strings_3}
str_c("prefix-", c("a", "b", "c"), "-suffix")
```

### Subsettings Strings


You can extract parts of a string using `str_sub()`. As well as the string, `str_sub()` takes start and end arguments which give the (inclusive) position of the substring:

```{r subsettings_strings_1}
x <- c("Apple", "Banana", "Pear")
str_sub(x, 1, 3)
str_sub(x, -3, -1)
```

Note that `str_sub()` won’t fail if the string is too short: it will just return as much as possible:

```{r subsettings_strings_2}
str_sub("a", 1, 5)
```
You can also use the assignment form of `str_sub()` to modify strings:

```{r subsettings_strings_3}
str_sub(x, 1, 1) <- str_to_lower(str_sub(x, 1, 1))
x
```

### Locales

```{r locales_1}
x <- c("apple", "eggplant", "banana")

str_sort(x, locale = "en")  # English
str_sort(x, locale = "haw") # Hawaiian
```

### Exercises

1. In code that doesn’t use stringr, you’ll often see `paste()` and `paste0()`. What’s the difference between the two functions? What stringr function are they equivalent to? How do the functions differ in their handling of NA?

```{r 2_1_1}

```


2. In your own words, describe the difference between the sep and collapse arguments to `str_c()`.

3. Use `str_length()` and `str_sub()` to extract the middle character from a string. What will you do if the string has an even number of characters?

```{r 2_2_1}

```

4. What does `str_wrap()` do? When might you want to use it?

5. What does `str_trim()` do? What’s the opposite of `str_trim()`?

6. Write a function that turns (e.g.) a vector `c("a", "b", "c")` into the string a, b, and c. Think carefully about what it should do if given a vector of length 0, 1, or 2.


## Matching patterns with regular expressions

```{r regex_1, echo=TRUE}
dot <- "\\."
writeLines(dot)
# swap with str_view
str_view(c("abc", "a.c", "bef"), "a\\.c")
```

### Exercises


1. Create regular expressions to find all words that:


Start with a vowel

```{r 3_1_1}
str_view(words, '^[aeiou]', match = TRUE)
```

contain only consonents

```{r 3_1_2}
str_view(words, '^[^aeiou]+$', match = TRUE)
```

End with ed, but not with eed

```{r 3_1_3}
str_view(words, '[^e]ed$', match = TRUE)
```

End with ing or ise.


```{r 3_1_4}
str_view(words, '(ing|ise)$', match = TRUE)
```

Have three or more vowels in a row.

```{r 3_1_5}
str_view(words, '[aeiou]{3,}', match=T)
```

start and end with the same character

```{r 3_1_6}
str_view(words, '^(\\w).*\\1$', match = TRUE)
```


Contain a repeated pair of letters (e.g. “church” contains “ch” repeated twice.)


```{r 3_1_7}
str_view(words, '([\\w]{2}).*\\1', match = TRUE)
```


2. Empirically verify the rule “i before e except after c”.

```{r 3_2_1}
str_view(words, '[^c]ie', match = TRUE)
```


3. Is “q” always followed by a “u”?

```{r 3_3_1}
str_view(words, 'q[^u]', match = TRUE)
```


4. describe what these expression will match


```{r 3_4_1}
str_view(words, '(.)(.)(.).*\\3\\2\\1', match = TRUE)
```


## Tools

### Detect Matches


```{r detect_1}
x <- c("apple", "banana", "pear")
str_detect(x, "e")
```


```{r detect_2}
# How many common words start with t?
sum(str_detect(words, "^t"))

# What proportion of common words end with a vowel?
mean(str_detect(words, "[aeiou]$"))
```

```{r detect_3}
# Find all words containing at least one vowel, and negate
no_vowels_1 <- !str_detect(words, "[aeiou]")
# Find all words consisting only of consonants (non-vowels)
no_vowels_2 <- str_detect(words, "^[^aeiou]+$")
# they should be the same
identical(no_vowels_1, no_vowels_2)
```


A common use of `str_detect()` is to select the elements that match a pattern. You can do this with logical subsetting, or the convenient `str_subset()` wrapper:

```{r detect_4}
# select elements
words[str_detect(words, "x$")]
# or easier
str_subset(words, "x$")
```

With dataframes use `filter`

```{r detect_5}
df <- tibble(
    word = words,
    i = seq_along(word)
)
df %>% 
    filter(str_detect(word, "x$"))
```


A variation on `str_detect()` is `str_count()`: rather than a simple yes or no, it tells you how many matches there are in a string:

```{r detect_6}
# On average, how many vowels per word?
mean(str_count(words, "[aeiou]"))
```

In dataframes this works well with `mutate`

```{r detect_7}
df %>% 
    mutate(
        vowels = str_count(word, "[aeiou]"),
        consonants = str_count(word, "[^aeiou]")
    )
```

### Extract Matches


```{r extract_1}
head(sentences)
```

find sentences that contain a colour

```{r extract_2}
colours <- c("red", "orange", "yellow", "green", "blue", "purple")
colour_match <- str_c(colours, collapse = "|")
colour_match
# sentences that have a colour
has_colour <- str_subset(sentences, colour_match)
has_colour
# extract the colours
matches <- str_extract(has_colour, colour_match)
head(matches)
```

Note that `str_extract()` only extracts the first match. We can see that most easily by first selecting all the sentences that have more than 1 match:


```{r extract_3}
more <- sentences[str_count(sentences, colour_match) > 1]
str_view_all(more, colour_match)
```



```{r extract_4}
str_extract(more, colour_match)
```



```{r extract_5}
str_extract_all(more, colour_match)
```


If you use `simplify = TRUE`, `str_extract_all()` will return a matrix with short matches expanded to the same length as the longest:

```{r extract_6}
str_extract_all(more, colour_match, simplify = TRUE)
x <- c("a", "a b", "a b c")
str_extract_all(x, "[a-z]", simplify = TRUE)
```

### Exercises

1. In the previous example, you might have noticed that the regular expression matched “flickered”, which is not a colour. Modify the regex to fix the problem.

```{r 4_1_1}
colour_match_word = str_c('\\b', colour_match, '\\b')
colour_match_word
str_view_all(
    sentences[str_count(sentences, colour_match_word) > 0],
    colour_match_word
)
```

### Grouped matches


```{r group_1}
noun <- "(a|the) ([^ ]+)"

has_noun <- sentences %>%
    str_subset(noun) %>%
    head(10)
has_noun
```

```{r group_2}
has_noun %>% 
    str_extract(noun)
```

Note that `str_extract()` gives us the complete match; `str_match()` gives each individual component. Instead of a character vector, it returns a matrix, with one column for the complete match followed by one column for each group:


```{r group_3}
has_noun %>% 
    str_match(noun)
```


```{r group_4}
all_matches <- has_noun %>% 
    str_extract_all(noun)
str(all_matches)

```


```{r group_5}
unlist(all_matches)
```


```{r group_6}
has_noun %>% 
    str_extract_all(noun, simplify = TRUE) %>%
    as_tibble_col()
```

```{r group_7}
tibble(sentence = sentences) %>% 
    tidyr::extract(
        sentence,
        c("article", "noun"),
        "(a|the) ([^ ]+)",
        remove = FALSE
    )
```

### Exercises

1. Find all words that come after a “number” like “one”, “two”, “three” etc. Pull out both the number and the word.

```{r 5_1_1}

```


2. Find all contractions. Separate out the pieces before and after the apostrophe.


```{r 5_2_1}

```

